---
title: "DIETFITS code AJCN figures"
author: "GK Fragiadakis"
date: "10/13/20"
output: html_document
---

## Intro:

Here is the code to generate the main figures in the manuscript. Note: orange diet is "low-fat", green diet is "low-carb", following the color scheme in the manuscript. 

Load required packages for analysis:

```{r}

library(tidyverse)
library(phyloseq)
library(siggenes)

```

Loading the data:

```{r}

# set as working directory of github repo's R folder using setwd() and/or confirm using getwd():
# setwd()

getwd()

nutrition_data <- readRDS("../data/nutrition_data.rds")
rare_ps <- readRDS("../data/DIETFITS_dada2_phyloseq_obj_rarefied.rds")

```

Plotting nutrition data (Figure 1B):

```{r}

ggplot(nutrition_data, aes(x=CollectionTime, y=weight, fill=Group)) +
  labs(title="Weight over time")+ 
  geom_boxplot() + scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  scale_color_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  theme_classic(base_size = 25)+scale_x_discrete(breaks=c(1,2,3,5), labels=c("BL", "3", "6","12")) + 
  labs(x="Timepoint (months)",y="Weight (kg)")

ggplot(nutrition_data,aes(x=CollectionTime,y=total_gfat,fill=Group)) + 
  geom_boxplot()+scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  scale_color_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+theme_classic(base_size = 25)+
  scale_x_discrete(breaks=c(1,2,3,5), labels=c("BL", "3", "6","12")) + 
  labs(title="Fat intake through time",x="Timepoint (months)",y="Total Fat (g)")

ggplot(nutrition_data,aes(x=CollectionTime,y=total_carb,fill=Group)) + 
  geom_boxplot()+scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  scale_color_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  theme_classic(base_size = 25) +scale_x_discrete(breaks=c(1,2,3,5),labels=c("BL", "3", "6","12")) + 
  labs(title="Carbohydrate intake through time",x="Timepoint (month)", y="Total Carbohydrate (g)")

ggplot(nutrition_data,aes(x=CollectionTime,y=total_dietary_fiber,fill=Group)) + 
  geom_boxplot()+scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  theme_classic()+scale_color_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat")) + 
  scale_x_discrete(breaks=c(1,2,3,5), labels=c("BL", "3", "6","12")) + 
  labs(title="Total change in dietary fiber through time", x="Timepoint",y="Total Dietary Fiber (g)")

ggplot(nutrition_data,aes(x=CollectionTime,y=total_protein,fill=Group)) + 
  geom_boxplot()+scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  theme_classic(base_size = 25) + scale_x_discrete(breaks=c(1,2,3,5),labels=c("Baseline", "3 months", "6 months","12 months")) +
  labs(title="Protein intake through time", x="Timepoint",y="Total Protein (g)")

ggplot(nutrition_data,aes(x=CollectionTime,y=energy,fill=Group)) + 
  geom_boxplot()+scale_fill_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  scale_color_manual(values=c("#99d594","#fc8d59"),labels=c("Low Carb", "Low Fat"))+
  theme_classic(base_size = 25) + scale_x_discrete(breaks=c(1,2,3,5),  labels=c("BL", "3", "6","12")) + 
  labs(title="Calorie intake through time", x="Timepoint (month)",y="Total kcal")

```

Removing duplicate timepoints (some patients had multiple samples per timepoint):

```{r}
metadata <- data.frame(rare_ps@sam_data) %>% select(old_record_id, CollectionTime)

Duplicate_timepoint <- duplicated(metadata)

rare_ps@sam_data$Duplicate_timepoint <- Duplicate_timepoint

# rarefied phyloseq object with duplicates removed:

diet_ps <- subset_samples(rare_ps, Duplicate_timepoint == FALSE)

```

## Taxa summaries

Calculates taxa composotion summaries at all levels ("Phylum", "Class", "Order", "Family", "Genus") and combines into a single data frame with metadata for analysis:

```{r}

# summarize taxa function: 

summarize_taxa <- function(phyloseq_obj, Rank, normalize = TRUE){

  taxa_df <- data.frame(phyloseq_obj@tax_table, stringsAsFactors = TRUE)
  otu_df <- data.frame(phyloseq_obj@otu_table)

  # renaming NA values from unassigned taxa as "Unassigned"
  levels(taxa_df[, Rank]) <- c(levels(taxa_df[, Rank]), "Unassigned")
  taxa_df[, Rank][is.na(taxa_df[, Rank])] <- "Unassigned"

  summary_data <- mat.or.vec(nr = dim(phyloseq_obj@otu_table)[2],
                             nc = length(levels(taxa_df[, Rank])))
  rownames(summary_data) <- colnames(phyloseq_obj@otu_table)
  colnames(summary_data) <- levels(taxa_df[, Rank])

  # ensure the components of the phyloseq object match
  if (identical(rownames(taxa_df), rownames(otu_df)) == FALSE){
    stop("Taxa and otu rownames don't match. Reformat phyloseq object.")
  }

  for (taxa in levels(taxa_df[, Rank])){
    taxa_counts <- apply(otu_df[taxa_df[, Rank] == taxa, ], 2, sum)
    summary_data[ , taxa] <- taxa_counts
  }

  if (normalize == TRUE){
    sums <- apply(summary_data, 1, sum)
    summary_data <- apply(summary_data, 2, function(x) x/sums)
  }

  return(summary_data)

}


ranks_of_interest <- c("Phylum", "Class", "Order", "Family", "Genus")

summary_list <- list()

for (rank in ranks_of_interest){
  
  summary_mat <- summarize_taxa(diet_ps, Rank = rank, normalize = TRUE)
  summary_df <- data.frame(SampleID = diet_ps@sam_data$SampleID, summary_mat) %>% 
    plyr::rename(c("Unassigned" = paste(rank, "_Unassigned")))
  summary_list[[rank]] <- summary_df
  
}

taxa_summaries <- Reduce(left_join, summary_list)

formatted_metadata <- data.frame(diet_ps@sam_data) %>% 
                            select(SampleID, Group, old_record_id, CollectionTime) %>% 
                            dplyr::rename(Timepoint = CollectionTime) %>%
                            dplyr::rename(Participant = old_record_id)

taxa_summaries <- left_join(formatted_metadata, taxa_summaries)

taxa_summaries$Timepoint <- as.numeric(taxa_summaries$Timepoint)

```

Running SAM to assess differences between timepoints: 

```{r}

## function to run SAM two-class paired 

SAM_obesity_dada2 <- function(data, diet, t1, t2){
  
  data_select_diet <- dplyr::filter(data, Group == diet)

    # restricting to the participants that have data for both of those time points
  selected_participants_t1 <- data_select_diet %>% dplyr::filter(Timepoint == t1) %>% select(Participant)
  selected_participants_t2 <- data_select_diet %>% dplyr::filter(Timepoint == t2) %>% select(Participant)
  selected_participants <- dplyr::intersect(selected_participants_t1, selected_participants_t2) %>% .[,1]
  
  
  print(paste("Chosen diet has ", length(selected_participants), " ppts", sep = ""))

  
  # check other diet, see how many have:
  
  data_other_diet <- dplyr::filter(data, Group != diet)

  selected_participants_t1 <- data_other_diet %>% dplyr::filter(Timepoint == t1) %>% select(Participant)
  selected_participants_t2 <- data_other_diet %>% dplyr::filter(Timepoint == t2) %>% select(Participant)
  selected_participants_other_diet <- dplyr::intersect(selected_participants_t1, selected_participants_t2)  %>% .[,1]
  
    print(paste("Other diet has ", length(selected_participants_other_diet), " ppts", sep = "")) 

  # restrict to minimum # of the 2 diets
    
    if (length(selected_participants) > length(selected_participants_other_diet)){
      set.seed(1)
 selected_participants <- sample(selected_participants, length(selected_participants_other_diet), replace = FALSE)
    }
    
  print(paste("Using ", length(selected_participants), " ppts", sep = ""))

# now setting up the SAM
  df <- data_select_diet %>% dplyr::filter(Participant %in% selected_participants) %>% 
    dplyr::filter(Timepoint == t1 | Timepoint == t2)

  
  sam_key <- df %>% select(Participant, Timepoint) %>% mutate(Sam_index = Timepoint)
  sam_key[sam_key$Sam_index == t1, "Sam_index"] <- -1
  sam_key[sam_key$Sam_index == t2, "Sam_index"] <- 1


  k = 1
  for (i in unique(sam_key$Participant)){
    sam_key[sam_key$Participant == i, "Participant_ID"] <- k
    k <- k + 1
  }

  sam_key <- sam_key %>% mutate(SAM_ID = Sam_index*Participant_ID) %>% select(Participant, Timepoint, SAM_ID)

    df <- left_join(sam_key, df)

  # Run Model

  y <- df[, "SAM_ID"]
  x <- df %>% select(-Participant, -Timepoint, -SAM_ID, -SampleID, -Group) %>% t()

  sam_model <- siggenes::sam(data = x, cl = y, gene.names = rownames(x))
  return(sam_model)
  
}


## Running the function:  cycle through both diets and all time points compared to baseline

taxa_SAM_results <- list()

for (diet_type in c("green", "orange")){
  
  for (time in c(2, 3, 4, 5)){
    
    condition <- paste(diet_type, "_1_", time, sep = "")
    model_result <- SAM_obesity_dada2(data = taxa_summaries, diet = diet_type, t1 = 1, t2 = time)
    taxa_SAM_results[[condition]] <- summary(model_result, 1.1)
    
  }
  
}

# Big changes 0 to 3 months, but not in the subsequent times. 
# q-value < 0.05

```


Confirmation that there are are not baseline differences between the two diet groups: 


```{r}

bl_df <- taxa_summaries %>% dplyr::filter(Timepoint == 1)
bl_df$Group <- recode(bl_df$Group, orange = 0, green = 1)

  y <- bl_df$Group
  x <- bl_df %>% select(-Participant, -Timepoint, -SampleID, -Group) %>% t()

  sam_model <- siggenes::sam(data = x, cl = y, gene.names = rownames(x))
  test <- siggenes::findDelta(sam_model, fdr = 0.05)

```


Looking at the three month result in each diet: 


```{r}

green_taxa <- names(taxa_SAM_results[["green_1_2"]]@row.sig.genes)
green_taxa

orange_taxa <- names(taxa_SAM_results[["orange_1_2"]]@row.sig.genes)
orange_taxa

overlapping_changes <- intersect(green_taxa, orange_taxa)
overlapping_changes
# Whole bacteroides lineage
# Whole parabacteroides lineage

plot_df <- gather(taxa_summaries, key = "Taxa", value = "Abundance", p__Actinobacteria:`Genus _Unassigned`) %>%
  dplyr::filter(Timepoint == 1 | Timepoint == 2)
#plot_df <- gather(taxa_summaries, key = "Taxa", value = "Abundance", p__Actinobacteria:`Genus _Unassigned`)
plot_df$Timepoint <- factor(plot_df$Timepoint)
```

Plot for Figure 4A: 

```{r}

overlap_df <- dplyr::filter(plot_df, Taxa %in% overlapping_changes)
overlap_df$Taxa <- factor(overlap_df$Taxa, levels = c("p__Bacteroidetes", "c__Bacteroidia", "o__Bacteroidales", "f__Bacteroidaceae", "g__Bacteroides", "f__Porphyromonadaceae", "g__Parabacteroides"))
ggplot(overlap_df, aes(x = Timepoint, y = Abundance)) + geom_boxplot(aes(fill = Group)) + facet_wrap(~Taxa, ncol = 5, scales = "free") + scale_fill_manual(values = c("#99d594","#fc8d59")) + theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        panel.border = element_rect(colour = "black"))

```

Plot for Figure 3A (Green diet/ low-carb changes only):

```{r}

green_only <- green_taxa[!(green_taxa %in% overlapping_changes)]
green_df <- dplyr::filter(plot_df, Taxa %in% green_only)
green_df$Taxa <- factor(green_df$Taxa, levels = c("c__Betaproteobacteria", "o__Burkholderiales", "f__Alcaligenaceae", "g__Sutterella", "c__Deltaproteobacteria", "o__Desulfovibrionales", "f__Desulfovibrionaceae", "g__Desulfovibrio", "g__Bilophila", "f__.Odoribacteraceae.", "g__Butyricimonas", "f__Rikenellaceae","g__Oscillospira", "g__Lachnospira" ))


green_taxa_descriptions <- read.csv("../data/green_taxa_descriptions.csv")
green_df_full <- left_join(green_df, green_taxa_descriptions)
green_df_full$Timepoint <- factor(green_df_full$Timepoint)
green_df_full$Level <- factor(green_df_full$Level, levels = c("Class","Order","Family","Genus"))
green_df_full$Phylum <- factor(green_df_full$Phylum, levels = c("Proteobacteria","Bacteroidetes", "Firmicutes"))


ggplot(green_df_full, aes(x = Timepoint, y = Abundance)) + geom_boxplot(aes(fill = Phylum)) + facet_wrap(~Taxa, ncol = 4, scales = "free") + scale_fill_manual(values = c("tomato","lightgoldenrod1", "skyblue2")) + theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        panel.border = element_rect(colour = "black"))
```

Plot for Figure 3B (Orange diet/ low-fat changes only):

```{r}

orange_only <- orange_taxa[!(orange_taxa %in% overlapping_changes)]
orange_df <- dplyr::filter(plot_df, Taxa %in% orange_only)
orange_df$Taxa <- factor(orange_df$Taxa, levels = c("p__Firmicutes", "c__Clostridia", "o__Clostridiales", "f__Lachnospiraceae", "g__Blautia" , "g__Dorea", "g__.Ruminococcus.", "p__Actinobacteria", "c__Actinobacteria", "o__Bifidobacteriales", "f__Bifidobacteriaceae", "g__Bifidobacterium"))

orange_taxa_descriptions <- read.csv("../data/orange_taxa_descriptions.csv")

# adding a superfluous plot so same dims as the green plot 
added_for_img <- dplyr::filter(green_df_full, Taxa == "g__Oscillospira")
orange_df_full <- left_join(orange_df, orange_taxa_descriptions)
orange_df_full$Timepoint <- factor(orange_df_full$Timepoint)
orange_df_full <- bind_rows(orange_df_full, added_for_img)
orange_df_full$Level <- factor(orange_df_full$Level, levels = c("Phylum", "Class","Order","Family","Genus"))
orange_df_full$Phylum <- factor(orange_df_full$Phylum, levels = c("Actinobacteria", "Firmicutes"))

orange_df_full <- orange_df_full %>% arrange(Phylum, Level)
orange_df_full$Taxa <- factor(orange_df_full$Taxa, levels = unique(orange_df_full$Taxa))
ggplot(orange_df_full, aes(x = Timepoint, y = Abundance)) + geom_boxplot(aes(fill = Phylum)) + facet_wrap(~Taxa, ncol = 4, scales = "free") + theme_classic() + scale_fill_manual(values = c("plum2", "skyblue2"))

ggplot(orange_df_full, aes(x = Timepoint, y = Abundance)) + geom_boxplot(aes(fill = Phylum)) + facet_wrap(~Taxa, ncol = 4, scales = "free") + scale_fill_manual(values = c("plum2", "skyblue2")) + theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        panel.border = element_rect(colour = "black"))

# note! Oscillospira is not one of the results! In there to force scale


```

Beta diversity (Figure 2):

```{r}

diet_ord_bray <- ordinate(diet_ps, "PCoA", "bray")

PC_df <- data.frame(diet_ord_bray$vectors[, 1:3], 
                    Participant = diet_ps@sam_data$old_record_id, 
                    Group = diet_ps@sam_data$Group,
                    Timepoint = diet_ps@sam_data$CollectionTime) %>%
        mutate(T2 = Timepoint == 2) %>% mutate(T1 = Timepoint == 1) %>% arrange(Axis.1) 

# reordering ppts
PC_df$Participant <- factor(PC_df$Participant, levels = unique(PC_df$Participant))


# Figure 2

ggplot(PC_df, aes(x = Axis.1, y = Participant, group = Participant)) + 
  geom_line(colour = "grey") + geom_point(aes(colour = T2, shape = T1)) +  
  facet_grid(Group ~., scales = "free_y") + theme_classic() + 
  scale_colour_manual(values = c("slategrey", "brown1")) + ggtitle("Bray Curtis over time") +
  scale_shape_manual(values=c(20, 1))

```

